---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Objective
To demonstrate a working knowledge of Literate (Statistical) Programming
using R Markdown 

## Loading and preprocessing the data
First thing we should do after reading data is try to understand it.

```{r}
activity <- read.csv("activity.csv")
str(activity)
summary(activity)
```

Here we can see there are 17,568 observations and 2304 of them are NA's or
about 13%.  
Also we know we will have to "fix" the date so it is in a usable format.  
The number of observations per day is complete (288), i.e.: there are no
days with missing records, although there are NA's.  
A question that will be answered later is: are missing values inside any 
given day or missing values correspond to complete days?  
It is somehow disturbing that the interval value goes up to 2355
(288*5 is 1440) and it can be explained by looking at the first 15 records.

```{r}
head(activity,15)
```

Ok! the interval is formed by four digits. The first two correspond to the
"hour" and the other two correspond to the minute, which actually is recorded
every five minutes (or twelve times in an hour).  
Records 0:12 correspond to hour 0, records 13:25 to hour 1, and so on.

Now we should convert dates to a date format.

```{r}
activity$date <- as.Date(as.character(activity$date), "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
First we determine the total number of steps each day, and obtain basic 
statistics.

```{r}
stepsbyday <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
summary(stepsbyday)
``` 

So we observe the mean is 9,354 and the median 10,400

Now we can answer the question previously stated. Which dates have 0 steps
recorded?
```{r}
stepsbyday[stepsbyday==0]
```

We can see 8 days have 0 steps recorded and it makes sense. It will be very
rare to have missing values inside a day. and just validating 8*288 is 2,304
or the total number of missing values.

An histogram of the steps by day follows: 
```{r}
hist(stepsbyday, breaks=10)
```   

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
